Quantifying Limits to Detection of Early Warning for Critical Transitions

### 2011-07-17 appendix/appendix.tex
some edits on appendix, finished synthesis of feedback from Marissa, Marcel, Alistair, Alex

\section{Model choice and model adequacy}\label{modelchoice}
Any approach is only as good as its underlying assumptions.  In this paper we have leveraged the machinery of likelihood-based approaches to both evaluate the performance of existing methods and provide a statistical test that more closely matches the assumptions of the basic theory.  For this approach to be useful, the data must still be well-approximated by the models used.  We have provided two different simple bifurcation models that could each represent the hypothesis of gradual decline in stability: linearized transcritical bifurcation and the linearized saddle-node bifurcation; Equations~\eqref{LTC},~\eqref{LSN}.  As they both represent a gradual loss of stability either can be used, but the test will be most powerful by using the one that best corresponds to the data.  While knowledge of the system alone can suggest this (such as in the case of~\citet{Drake2010}), it is straight forward to compare the models against each other using the same likelihood framework, as illustrated in Figure~\ref{fig:modelchoice} for several empirical examples.

\begin{figure}[ht]
\begin{center}
\includegraphics[width=.25\textwidth]{deut_modelchoice.png}
\includegraphics[width=.25\textwidth]{caco3_modelchoice.png}
\end{center}
\caption{Model }
\label{fig:modelchoice}
\end{figure}

@@

% Choosing between models

>>>>

\section{On a hypothesis-testing framework}
It is possible to frame the question of sensitivity, reliability, and adequate data in the language of hypothesis testing. As this introduces the unnecessary nuisance parameter, the statistical significance criterion, it is not presented in the text.  In the hypothesis testing framework, a false positive is a Type I error, which is defined relative to some arbitrary statistical significance criterion, most commonly 0.05.  By changing this criterion, one can increase or decrease the probability of the Type I error at the cost of decreasing or increasing the Type II error, also defined relative to this criterion.

The work by~\citet{Dakos2008} is one of the few commendable examples that attempts a careful quantification of the statistical significance of its approaches.  While the $p$-values presented in the main text are based on parametric assumptions not met by the data, the rather different $p$ values presented in appendix table S3 are constructed with more reasonable null hypotheses (in particular, H$_0$ 2 and H$_0$ 3 are very similar to our stable system, though H$_0$ 1 has lost the correlation structure expected in any time-series).  The plots in Fig. 3 confirm our finding that almost any value of Kendall's $\tau$ is possible under the null hypothesis, and the $p$-values in table S3 show that the detection method is not significant in any of the time series that had fewer than 400 points.  Our approach replicate these findings for the $\tau$-statistic.  We were glad to find that our likelihood-based approach does not have this difficulty, easily identifying the warning signal in these shorter time series, while also reliably avoiding false alarms.

The approach taken by~\citet{Dakos2008} does not present the corresponding analysis of power or Type II error.  As statistical significance could be achieved with a method that selects less than one in every twenty data sets at random, it is impossible to know whether the approach is actually detecting a signal or simply a random pattern in the noise without a test of power.  Fig. S3, highlights the importance of so doing -- the null distribution is almost uniform across all possible values of $\tau$.  The distribution of $\tau$ under the test distribution will not be able to avoid at least some overlap on this bounded domain.  The reality is worse, we find it overlaps almost completely.  We were glad to see that in estimating the distribution of $\tau$ for very large data sets, it does tighten around zero, while the distribution expected under a true positive condenses around more positive values, indicating that there is at least some discriminative ability in the test.

The essential features of these overlapping distributions are easily captured in the ROC curves, which provides a natural language to discuss the issues of uncertainty -- such as ``accepting a higher false-alarm rate in order to avoid a catastrophic shift'' rather than ``using a more liberal threshold for $p$-values in order to avoid costly Type II error.''  %A statistical test is meaningless without at least some notion of power.  As it is perhaps the linguistics of hypothesis testing as much as the quantitative difficulty of the analysis that has led this step to be entirely ignored (though we admit some surprise at the extent to which the question of significance or false alarms has also been ignored), we hope

\section{Model choice and model adequacy}\label{modelchoice}
Any approach is only as good as its underlying assumptions.  In this paper we have leveraged the machinery of likelihood-based approaches to both evaluate the performance of existing methods and provide a statistical test that more closely matches the assumptions of the basic theory.  For this approach to be useful, the data must still be well-approximated by the models used.  We have provided two different simple bifurcation models that could each represent the hypothesis of gradual decline in stability: linearized transcritical bifurcation and the linearized saddle-node bifurcation; Equations~\eqref{LTC},~\eqref{LSN}.  As they both represent a gradual loss of stability either can be used, but the test will be most powerful by using the one that best corresponds to the data.  While knowledge of the system alone can suggest this (such as in the case of~\citet{Drake2010}), it is straight forward to compare the models against each other using the same likelihood framework.

@@
